Detecting the impact of land cover change on observed rainfall.

### 2019-03-14 Manuscripts/PeerJRainfall_Landuse/Rainfall_Landuse.Rmd
paper for submission

---
title: Detecting the impact of land cover change on observed rainfall.
preprint: false
author:
  - name: Chun X. Liang
    affiliation: 1
  - name: Floris F. van Ogtrop
    affiliation: 1
  - name: R. Willem Vervoort
    affiliation: 1
    corresponding: true
    email: willem.vervoort@sydney.edu.au
affiliation:
  - code: 1
    address: Sydney Institute of Agriculture, The University of Sydney, NSW 2006
abstract: >
 Analysis of observational data to identify relationships between rainfall and land cover change are difficult due to multiple environmental factors that cannot be strictly controlled. In this study we present a methodology to investigate the relationship using statistical methods on data from best available sources at two sites in Australia. Gridded data of rainfall and tree cover were used as spatially corresponding local conditions. Large scale effects were represented by climatic indicators, such as SOI and IOD. Regression analysis and step trend tests were used to assess the effect of abrupt land surface intervention. 
  At a Queensland site, significant tree cover change between 2002 - 2005 did not result in strong statistically significant precipitation changes. On the other hand, results from a bushfire affected NSW/VIC region suggests significant changes in the rainfall. This indicates the method works better when a abrupt change in the data can be clearly identified. The results from the step trend test implied a positive relationship between the tree cover and the rainfall at 0.1 significance level in both locations in data up to 2009. However, high rainfall variability and possible regrowth meant that no signifcant changes were observed in alonger time sereies to 2015. 
header-includes: >
  \usepackage{lipsum}
  \usepackage{textcomp}
bibliography: reflist_chap3.bib
date: "`r Sys.Date()`"
output: 
  rticles::peerj_article:
    base_format: bookdown::pdf_document2 # bookdown::pdf_document2 # for using \@ref()
---
```{r packages, echo=F, warnings=F, message=F, eval=T}
require(knitr)
require(bookdown)
require(pander)
library(fields)
library(graphics)
library(maptools)
library(plotKML)
library(raster)
library(reshape)
library(rgeos)
library(rgdal)
library(RColorBrewer)
library(tidyverse)
require(lubridate)
library(deseasonalize)
library(mgcv)
library(ncdf4)
require(gridExtra)
#library(nnet)
```

@@

Land use and land cover changes can lead to changes in the local climate. Empirical and modelling studies have found that cloud types and rainfall are correlated to large scale vegetation cover changes, such as deforestation in the Amazon and in the Sahel [@Chagnon2005;@Pinto2009; @Wang2009; @Mei2010; @kucharski_further_2013; @pitman_scale_2016] and afforestation in south Israel [@Otterman1990; @Ben-Gai1998]. Using airborne measurement in Western Australia, @Junkermann2009 showed a significantly higher level of aerosols over an agricultural area compared to an adjacent natural vegetation. They suggested that a modification of aerosol concentrations due to deforestation could have contributed to a reduction of local rainfall, as more, but smaller rain droplets were observed. @Nair2011 reported from the Bunny Fence Experiment in Western Australia that local land use change altered the synoptic west coast trough dynamics and surface roughness, and this resulted in an observed rainfall decrease. Maximum temperatures were also found to be sensitive to land cover change in eastern Australia [@McAlpine2007].

@@

Therefore, the aim of this study is to use a statistical approach on rainfall data at regional scales to investigate the cause and effect relationship between land cover change and local rainfall, which is demonstrated in many modelling studies. More specifically, we hypothesize that a step change on the land cover on the surface will cause a step change in the rainfall. To demonstrate this we study changes in observed rainfall over time at a Queensland and NSW/Victoria location where there are possible step changes in land cover change due to land clearing and bush fires. The methodology uses statistical approaches to identify changes in rainfall, which are subsequently associated with land cover change through spatial comparison.

@@

In Australia, significant tree cover change has mainly occurred in the north east  and south east of the continent,  as well as in the southwest of Western Australia. According to the National Dynamic Land Cover Dataset (DLCD) [@Lymburner2010], most of these areas have experienced decreases in the Enhanced Vegetation Index (EVI) post 2000 aws derived from satellite data. As an index for vegetation greenness, the decreasing values indicate lower biomass over time in the tree cover regions. The possible EVI reduction might be due to land clearing, bush fires or drought.

@@

Two regions were selected where significant tree cover change since 2000 was reported. The first region is located in south central Queensland partly covering the north of the Murray Darling Basin (MDB) (site 1 in Figure \@ref(fig:selreg)). High rates of land clearing have been reported in this region during the early 2000s [@SLATS2004]. The second study region is located at the border of New South Wales and Victoria, and includes the Snowy Mountain ranges (site 2 in Figure \@ref(fig:selreg)). Severe bush fires occurred in this area  in early 2003 (see Figure \@ref(fig:figure2bushfire)). The 2003 bush fires were the largest and the worst in the last 60 years [@Fire2011]. Two thirds of Kosciuszko national park was heavily burned and regrowth was reported to be slow due to drought and cold conditions [@ABC2003], and the type of species in this region. However, in the longer term, after an early high transpiration period a recovery of pre-fire evapotranspiration would be expected [@Kuzcera1987]. For the purpose of this study, significant tree cover loss has happened in both study areas in the last decade, either permanently or temporarily.

```{r figure2bushfire, fig.cap = "Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels. The map shows the large area in the Kosciuszko national park that has been burned. Some locations in the southwest of teh Australian Capital Territory (ACT) have also experienced intensive bushfires.", out.width="90%", echo=F}
include_graphics("figures/bushfire_nswvic.jpg")
```

@@

Data {#Data}
==============
Several land surface data sets were used in this study. The main one was the MOD44B product Global Vegetation Continuous Field data set (version 5). This data set provides estimates of percent tree cover (percentage of ground surface covered by trees) at a grid resolution of 250 m [@Townshend2011], which is finer then the earlier mentioned burned product MCD45A1. The data set is available on an annual time interval for the study period of 2000 - 2015. The tree cover data was produced from 16-day Terra MODIS Land Surface Reflectance data and Land Surface Temperature [@Townshend2011]. The National Dynamic Land Cover Dataset (DLCD) [@Lymburner2010] from the Australian Collaborative Land Use Mapping Program (ACLUMP) was used to verify the trend of vegetation cover change calculated from the previous data set. This data set, developed by Geoscience Australia and Australian Bureau of Agricultural and Resource Economics and Sciences (ABARES), is the first nationally consistent and thematically comprehensive land cover reference for Australia. The DLCD is based on the 16-day Enhanced Vegetation Index (EVI), again from the MODIS satellite, between April 2000 and April 2015. It also has a grid resolution of 250 m. The data set provides information on the final land cover types (as in 2015) and estimated trend of EVI statistics (annual mean, maximum and minimum).

@@

### Regression model {#reg_model}

As highlighted in the introduction, the Australian climate is influenced by sea surface temperatures in the tropical Pacific and Indian Oceans, as well as pressure systems in the Southern Ocean [@BoM2012].  @Risbey2009 compared five large-scale drivers, including ENSO (measured by SOI and the Tropical Pacific Sea surface temperatures (SSTs)), IOD, SAM, MJO (Madden-Julian oscillation) and blocking, in relation to Australian rainfall variability. The MJO is a large scale eastward-propagating wave-like disturbance located around equatorial latitudes [@Risbey2009]. They identified SOI as the most important index among all climate indices tested for broad parts of Australia (including QLD and NSW/VIC) in almost any season. In this study, four climate indices were selected from the main climatic indicators (see section [Data]) and used as the explanatory variables in the model for each study region. A further complicating factor is the influence of the "millenium drought" over the study period and in particular the change to wet conditions in 2010 - 2011 [@vanDijk2013]. Therefore, the spatially averaged monthly rainfall in the Murray Darling Basin (MDB, downloaded from [the Bureau of Meteorology](http://www.bom.gov.au/web01/ncc/www/cli_chg/timeseries/rain/allmonths/mdb/latest.txt)) was used to explain the year-on-year variation in the rainfall in the regions. As climate alone are unlikely able to fully explain the temporal variation in rainfall in the regions, including the drought [@vanDijk2013; @Westra2010], this variable was included. Since both regions at least partly overlap with the MDB, the average rainfall across was assumed to be a useful explaining variable.

@@

Correlations between rainfall and each climate index were analysed. Rainfall in each study region was first deseasonalised and detrended using the seasonal decomposition function `ds` in the package `deseasonalise` in R [@Rstats2018]. Using detrended data gives a better indication of the underlying correlation compared to the correlation between trends in the data [@Smith2012]. 
The cross-correlations between the deseasonalised and detrended rainfall and the climatic indices were tested using the Pearson's product moment correlation method, assuming the relationships are linear. Although the optimal technique for exploring the correlation with each index could be different as described in @Risbey2009, the Pearson's method was applied to all indices for consistency. Because the PDO describes the multi-decadal SST with lower frequency [@MacDonald2005; @Zanchettin2008; @Kamruzzaman2011], instead of 37-year rainfall data, a longer period (115 years, from 1900 to 2015) was used to estimate the correlation with PDO, up to lag 24. For the other indices, the 37-year data was used.

@@

Based on the correlation between the climatic indices and rainfall in the regions (as shown in Figure \@ref(fig:cor-rain-qld) and Figure \@ref(fig:cor-rain-nsw)), it can be concluded that:
\begin{itemize}
  \setlength{\itemsep}{0cm}
  \setlength{\parskip}{0cm}
  \item In QLD, the correlation between rainfall and SOI at zero time lags is the strongest across all indices, outweighing the other ENSO indicators. IOD and PDO have a weak influence in QLD.
  \item In NSW/VIC, again the SOI has the strongest correlation with rainfall, followed by the IOD. For both, the strongest correlations occur at the zero time lags. 
  \item in both cases PDO had the weakest correlations, and this factor was not further considered as a predictor.
  \item in general, for the better correlated indices, strongest correlations occured at zero time lags.
\end{itemize}
The above findings are consistent with previous studies. Although some indices are serially correlated with rainfall up to several months, the lag zero events have the greatest correlation coefficients. Using multiple climatic index series was generally found most useful in rainfall prediction [e.g. @Risbey2009; @Kamruzzaman2011].

@@

We assumed all the factors are additive smooth components in determining rainfall following @Kamruzzaman2011. In this case, the rainfall model is a generalised additive model (GAM) [@Hastie1986] with a log link function `g()` and assuming the residuals are gamma distributed (see Figure \@ref(fig:hist-rain)). This means all predictors are modelled as smooth functions, in this case using the shrinkage version of the cubic regression splines [@Wood2011]. All splines were limited to 3 knots in flexibility [@Wood2011], to reduce the risk of overfitting.
\vspace{0.5cm}
\begin{equation}
\begin{array}{lll}
g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{MDB_{monthlyRain}} + \\ &s_2(\mathbf{SOI}) + s_3(\mathbf{IOD}) + \\                &s_4(\mathbf{Nino3.4}) + s_5(\mathbf{Nino4}) + s_6(\mathbf{Season}) + \\
           &\beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
\end{array}
(\#eq:model)
\end{equation}
The bold letters represent the time series vectors.  The region is indicated by $r$, while $\beta_u$ ($_u$=0, 1) are the fitted coefficients in the model. $s_v$ ($_v$=1, 2, 3,...) are the smooth penalized cubic regression spline functions on the climatic indices and the season. MDB_{monthlyRain} is the spatially averaged monthly rainfall across the Murray Darling Basin. Apart from dropping PDO as a predictor, all other climate indices were included, allowing the model to select the appropriate predictors.

@@

One of the difficulties is to point an exact time to the changes in the vegetation cover in the two regions. In the QLD region, no exact time can be assigned to the land clearing. According to the SLATS reports, the most substantial clearing occurred between 2003 - 2004 . However, the information on the change in type of land cover during the time period is missing. Therefore, four scenarios were initially tested in the analysis (**see supplementary material**). In the NSW/VIC region, severe bush fires were reported in early January 2003. Hence the "tree" cover state was up to December 2002 then it was changed to "removed" state from January 2003. As a starting date, the regression model was run from 1979 for both regions.

@@

To support the regression analysis, a Mann-Whitney Rank-Sum step trend test was used to detect changes in rainfall as a result of vegetation cover change. This specific nonparametric statistical test was modified from the Mann-Whitney U test by @Hirsch1985 and can identify a step change in data which is cross-correlated. The gridded rainfall dataset used in this study has a high spatial correlation between neighbouring pixels due to the underlying interpolation method. The advantages of using the Rank-Sum test are: (1) it does not depend on assumptions of the data distribution; (2) it is not restricted to datasets with no missing data; (3) it is robust and not as easily influenced by outliers and negative numbers [@Hirsch1985].

@@

```{r exampleRank, echo=F}
table <- data_frame(Year = c(1998 , 1999 , 2000 , 2001 ,
                             2002 , 2005 , 2006),
                    `Rainfall residuals` =c(-0.3, -60.9, -16.1,
                                                      -71.7, 111.1,
                                             -7.2, -60.5),
                    Rank= c(6, 2, 4,
                                                               1, 7, 5, 3))
colnames(table)[3] <- paste("Rank","$R'_{1k}$")
pander(table, caption="Example of ranking rainfall residuals",
             style="rmarkdown")

@@

The before and after period in the data formed two groups of samples. The split point of the two periods was based on the timing of the vegetation cover changes. In the QLD region, changes occurred anytime during 2003 and 2004. In contrast to the previous method, the time period covering the land cover change  was excluded, as the nonparametric test allows missing data. @Hirsch1985 also pointed out that the power of the test is higher if the data of the change period is ignored. Hence 2003 and 2004 were excluded from the analysis. As a result, the after-change period was 2005 - 2015 for the Queensland location.

In the case of NSW/VIC, the bushfires broke out in early January 2003. The change was within a relatively short period of the year. Therefore the after change period in this region still started in January 2003. Following @Hirsch1985, the before period was set to five years (1998 - 2002) in both regions. The length of the after period is also difficult for the NSW/VIC region, as the regrowth would at some point have impacted the local effects.

@@

Instead of completing the Mann-Whitney U-test, @Hirsch1985 applied a standard normal Z test to the rank-sum statistics. As highlighted, this modified test accounts for serial and cross correlation in the data. In the case here, the deseasonalised and detrended data shows little autocorrelation in the time series but possesses strong cross correlation between neighbouring pixels, i.e. $R>0.99$.

@@

```{r (ref:Zscore), echo=F}
table2 <- data_frame(" "= c("Z' $>$ 0 & rainfall decreases after change",
    "Z' $<$ 0 & rainfall increases after change",
    "Z' = 0 & rainfall does not change"))
pander(table2, caption="The interpretation of Z' score in the step trend test", style="rmarkdown")
```

@@

Generally the regression model only explains a limited amount of the rainfall variability (Figure \@reg(fig:rsq_model). The model in Equation \@ref(eq:model2) accounts for an average around 33\% of the variation in the NSW/VIC region and about 50% of the variation in the QLD region as indicated by the adjusted $r^2$. The adjusted $r^2$ is the coefficient of determination, a measurement of the amount of variability predicted by the model adjusting for the number of explanatory terms. The residual analysis shows that the assumptions of the regression model are generally met (Figure \@ref(fig:residuals). The standardised residual plots, however, show some funnelling for the NSW/VIC regions, suggesting non constant variance. The residual patterns are consistent for all pixels within each region.

```{r rsq_model, fig.cap = "The distribution of the variance explained (adjusted $r^2$ ) by the full regression model", echo=F, out.width="90%"}
include_graphics("figures/rsq_fullmodel.jpg")

@@

@@

```{r rsq, fig.cap = "The performance of the regression model if rainfall is only modelled by the climate drivers. It shows the percentage of rainfall variability that can be explained by the climate drivers for the Qld and NSW/VIC region", echo=F, out.width="90%"}
include_graphics("figures/soi_explhistogram.jpg")

@@

The spatial step trend test based on the regression residuals without the landcover variable generates Z' scores. The results for the Qld region (left) and NSWVIC region (right)  are shown in Figure \@ref(fig:steptest30). This figure provides two types of information: the sign and the significance level. The sign indicates the direction of the step change in the residuals of the regression, as listed in Table \@ref(Zscore). In both regions there are areas of positive Z' values which imply a decrease in rainfall, but this is stronger in the NSW/VIC area than in the QLD area. For the NSW/VIC area there appears to be a reasonable relationship between the locations where changes in tree cover are observed (Figure \@ref(fig:tc_trend)) and the patterns in the Z' score. However, there is not necessarily a direct relationship as movement of air masses could mean that actual changes of rainfall are observed close by, but not necessarily exactly at areas with changes of landcover.
There is once again a difference between the left panel and the right panel, indicating only a few significant of Z' scores in the QLD region wich matches the earlier significance in the landcover variable in the full regresion. In the QLD region, only 0.2\% of the pixels obtained a positive Z' score with p $<$ 0.1. In the NSW/VIC region 3.7\% of pixels have a positive Z' score with p $<$ 0.1. In general it is only a small proportion of both study regions.

```{r meandiff, fig.cap ="Boxplots of annual rainfall residuals (estimated based on Equation 2 before and after the land cover intervention during 1979 - 2015 in the study regions. On average, the after period has a significantly lower annual rainfall residual in NSW/VIC, but a signifcantly higher annual rainfall residual in the Qld study area", echo=F, out.width="90%"}

@@

The rainfall regression residuals without the landcover variable for the two regions (before-change since 1979 and after-change) were also compared (Figure \@ref(fig:meandiff)) using a simple t-test. From the boxplots it is difficult to see that there were significantly different ($p < 0.05$) mean rainfall residual values between the "before" and "after" periods in both regions.  For the Queensland locations, the mean monthly rainfall residuals were slightly lower  ($p < 0.05$) after the change in landcover ($~ 0.45 mm/month$). For the NSW/VIC locations there was larger decrease in the mean monthly rainfall residual ($p < 0.05$) after the change($~ 1.5 mm/month$).

>>>>

---
title: Detecting the impact of land cover change on observed rainfall.
preprint: false
author:
  - name: Chun X. Liang
    affiliation: 1
  - name: Floris F. van Ogtrop
    affiliation: 1
  - name: R. Willem Vervoort
    affiliation: 1
    corresponding: true
    email: willem.vervoort@sydney.edu.au
affiliation:
  - code: 1
    address: Sydney Institute of Agriculture, The University of Sydney, NSW 2006
abstract: >
 Analysis of observational data to pinpoint impact of land cover change on local rainfall are difficult due to multiple environmental factors that cannot be strictly controlled. In this study we use a statistical approach to identify the relationship between removal of tree cover and rainfall with data from best available sources for two large areas in Australia. Gridded rainfall data between 1979 and 2015 was used for the areas, while large scale (exogenous) effects were represented by mean rainfall across a much larger area and climatic indicators, such as SOI and IOD. Both general additive modelling and step trend tests were used for the analysis. 
  For a region in south central Queensland, the reported change in tree clearing between 2002 - 2005 did not result in strong statistically significant precipitation changes. On the other hand, results from a bushfire affected region on the border of New SOuth Wales and Victoria suggests significant changes in the rainfall due to changes in tree cover. This indicates the method works better when a abrupt change in the data can be clearly identified. The results from the step trend test also mainly identified a positive relationship between the tree cover and the rainfall at p < 0.1 at the NSW/Victoria region. High rainfall variability and possible regrowth could have impacted the results in the Queensland region.
header-includes: >
  \usepackage{lipsum}
  \usepackage{textcomp, rotating}
bibliography: reflist_chap3.bib
date: "`r Sys.Date()`"
output: 
  rticles::peerj_article:
    base_format: bookdown::pdf_document2 # bookdown::pdf_document2 # for using \@ref()
---
```{r packages, echo=F, warnings=F, message=F, eval=T}
suppressWarnings(require(knitr))
suppressWarnings(require(bookdown))
suppressWarnings(require(pander))
suppressWarnings(library(fields))
suppressWarnings(library(graphics))
suppressWarnings(library(maptools))
suppressWarnings(library(plotKML))
suppressWarnings(library(raster))
suppressWarnings(library(reshape))
suppressWarnings(library(rgeos))
suppressWarnings(library(rgdal))
suppressWarnings(library(RColorBrewer))
suppressWarnings(library(tidyverse))
suppressWarnings(require(lubridate))
suppressWarnings(library(deseasonalize))
suppressWarnings(library(mgcv))
suppressWarnings(library(ncdf4))
suppressWarnings(require(gridExtra))
suppressWarnings(require(kableExtra))
#library(nnet)
```

@@

Land use and land cover changes can lead to changes in the local climate. Empirical and modelling studies have found that cloud types and rainfall are correlated to large scale vegetation cover changes, such as deforestation in the Amazon and in the Sahel [@Chagnon2005;@Pinto2009; @Wang2009; @Mei2010; @kucharski_further_2013; @pitman_scale_2016] and afforestation in south Israel [@Otterman1990; @Ben-Gai1998]. Using airborne measurements in Western Australia, @Junkermann2009 showed a significantly higher level of aerosols over an agricultural area compared to an adjacent area with natural vegetation. They suggested that a modification of aerosol concentrations due to deforestation could have contributed to a reduction of local rainfall, as more, but smaller rain droplets were observed. @Nair2011 reported from the Bunny Fence Experiment in Western Australia that local land use change altered the synoptic west coast trough dynamics and surface roughness, and this resulted in an observed rainfall decrease. Maximum temperatures were also found to be sensitive to land cover change in eastern Australia [@McAlpine2007].

@@

Therefore, the aim of this study is to use a statistical approach on rainfall data at regional scales to investigate the cause and effect relationship between land cover change and local rainfall, which is demonstrated in many modelling studies. More specifically, we hypothesize that a step change in land cover on the surface will cause a step change in the rainfall. To demonstrate this we study changes in observed rainfall over time at a Queensland and NSW/Victoria location where there are possible step changes in land cover change due to land clearing and bush fires. The methodology uses statistical approaches to identify changes in rainfall, which are subsequently associated with land cover change through spatial comparison.

@@

In Australia, significant tree cover change has mainly occurred in the north east  and south east of the continent,  as well as in the southwest of Western Australia. According to the National Dynamic Land Cover Dataset (DLCD) [@Lymburner2010], most of these areas have experienced decreases in the Enhanced Vegetation Index (EVI) post 2000, as derived from satellite data. As an index for vegetation greenness, the decreasing values of EVI indicate lower biomass over time in the tree cover regions. The possible EVI reduction might be due to land clearing, bush fires or drought.

@@

Two regions were selected where significant tree cover change since 2000 was reported. The first region is located in south central Queensland (QLD) partly covering the north of the Murray Darling Basin (MDB) (site 1 in Figure \@ref(fig:selreg)). High rates of land clearing have been reported in this region during the early 2000s [@SLATS2004]. The second study region is located at the border of New South Wales and Victoria (NSW/VIC), and includes the Snowy Mountain ranges (site 2 in Figure \@ref(fig:selreg)). Severe bush fires occurred in this area  in early 2003 (see Figure \@ref(fig:bushfire)). The 2003 bush fires were the largest and the worst in the last 60 years [@Fire2011]. Two thirds of Kosciuszko national park was heavily burned and regrowth was reported to be slow due to drought and cold conditions [@ABC2003], and the type of species in this region. However, in the longer term, after an early high transpiration period a recovery of pre-fire evapotranspiration would be expected [@Kuzcera1987]. For the purpose of this study, significant tree cover loss has happened in both study areas in the last decade, either permanently or temporarily.

```{r bushfire, fig.cap = "Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels. The map shows the large area in the Kosciuszko national park that has been burned. Some locations in the southwest of the Australian Capital Territory (ACT) have also experienced intensive bushfires.", out.width="90%", echo=F}
include_graphics("figures/bushfire_nswvic.jpg")
```

@@

Data {#Data}
==============
Several land surface data sets were used in this study. The main one was the MOD44B product Global Vegetation Continuous Field data set (version 5). This data set provides estimates of percent tree cover (percentage of ground surface covered by trees) at a grid resolution of 250 m [@Townshend2011], which is finer than the earlier mentioned burned product MCD45A1. The data set is available on an annual time interval for the study period of 2000 - 2015. The tree cover data was produced from 16-day Terra MODIS Land Surface Reflectance data and Land Surface Temperature [@Townshend2011]. The National Dynamic Land Cover Dataset (DLCD) [@Lymburner2010] from the Australian Collaborative Land Use Mapping Program (ACLUMP) was used to verify the trend of vegetation cover change calculated from the previous data set. This data set, developed by Geoscience Australia and Australian Bureau of Agricultural and Resource Economics and Sciences (ABARES), is the first nationally consistent and thematically comprehensive land cover reference for Australia. The DLCD is based on the 16-day Enhanced Vegetation Index (EVI), again from the MODIS satellite, between April 2000 and April 2015. It also has a grid resolution of 250 m. The data set provides information on the final land cover types (as in 2015) and estimated trend of EVI statistics (annual mean, maximum and minimum).

@@

## Regression model {#reg_model}

As highlighted in the introduction, the Australian climate is influenced by sea surface temperatures in the tropical Pacific and Indian Oceans, as well as pressure systems in the Southern Ocean [@BoM2012].  @Risbey2009 compared five large-scale drivers, including ENSO (measured by SOI and the Tropical Pacific Sea surface temperatures (SSTs)), IOD, SAM, MJO (Madden-Julian oscillation) and blocking, in relation to Australian rainfall variability. The MJO is a large scale eastward-propagating wave-like disturbance located around equatorial latitudes [@Risbey2009]. They identified SOI as the most important index among all climate indices tested for broad parts of Australia (including QLD and NSW/VIC) in almost any season. In this study, four climate indices were selected from the main climatic indicators (see section [Data]) and used as the explanatory variables in the model for each study region. A further complicating factor is the influence of the "millenium drought" over the study period and in particular the change to wet conditions in 2010 - 2011 [@vanDijk2013]. Therefore, the spatially averaged monthly rainfall in the Murray Darling Basin (MDB, downloaded from [the Bureau of Meteorology](http://www.bom.gov.au/web01/ncc/www/cli_chg/timeseries/rain/allmonths/mdb/latest.txt)) was used to explain the year-on-year variation in the rainfall in the regions. As climate alone are unlikely able to fully explain the temporal variation in rainfall in the regions, including the drought [@vanDijk2013; @Westra2010], this variable was included. Since both regions at least partly overlap with the MDB, the average rainfall for the entire basin was assumed to be a useful explaining variable.

@@

Correlations between rainfall and each climate index were analysed. Rainfall in each study region was first deseasonalised and detrended using the seasonal decomposition function `ds` in the package `deseasonalise` in R [@Rstats2018]. Using detrended data gives a better indication of the underlying correlation by removing the underlying correlation in the data [@Smith2012]. 
The cross-correlations between the deseasonalised and detrended rainfall and the climatic indices were tested using the Pearson's product moment correlation method, assuming the relationships are linear. Although the optimal technique for exploring the correlation with each index could be different as described in @Risbey2009, the Pearson's method was applied to all indices for consistency. Because the PDO describes the multi-decadal SST with lower frequency [@MacDonald2005; @Zanchettin2008; @Kamruzzaman2011], instead of 37-year rainfall data, a longer period (115 years, from 1900 to 2015) was used to estimate the correlation with PDO, up to lag 24. For the other indices, the 37-year data was used.

@@

Based on the correlation between the climatic indices and rainfall in the regions (as shown in Figure \@ref(fig:cor-rain-qld) and Figure \@ref(fig:cor-rain-nsw)), it can be concluded that:
\begin{itemize}
  \setlength{\itemsep}{0cm}
  \setlength{\parskip}{0cm}
  \item In QLD, the correlation between rainfall and SOI at zero time lags is the strongest across all indices, outweighing the other ENSO indicators. IOD and PDO have a weak influence in QLD.
  \item In NSW/VIC, again the SOI has the strongest correlation with rainfall, followed by the IOD. For both, the strongest correlations occur at the zero time lags. 
  \item In both cases PDO had the weakest correlations, and this factor was not further considered as a predictor.
  \item In general, for the better correlated indices, strongest correlations occured at zero time lags.
\end{itemize}
The above findings are consistent with previous studies. Although some indices are serially correlated with rainfall up to several months, the lag zero events have the greatest correlation coefficients. Using multiple climatic index series was generally found most useful in rainfall prediction [e.g. @Risbey2009; @Kamruzzaman2011].

@@

We assumed all the factors are additive smooth components in determining rainfall following @Kamruzzaman2011. In this case, the rainfall model is a generalised additive model (GAM) [@Hastie1986] with a log link function `g()` and assuming the residuals are gamma distributed (see Figure \@ref(fig:hist-rain)). This means all predictors are modelled as smooth functions, in this case using the shrinkage version of the cubic regression splines [@Wood2011]. All splines were limited to 3 knots in flexibility [@Wood2011], to reduce the risk of overfitting.
\vspace{0.5cm}
\begin{equation}
\begin{array}{lll}
g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{MDB_{monthlyRain}} + s_2(\mathbf{SOI}) + s_3(\mathbf{IOD}) + \\
&s_4(\mathbf{Nino3.4}) + s_5(\mathbf{Nino4}) + s_6(\mathbf{Season}) + \\
           &\beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
\end{array}
(\#eq:model)
\end{equation}
The bold letters represent the time series vectors.  The region is indicated by $r$, while $\beta_u$ ($_u$=0, 1) are the fitted coefficients in the model. $s_v$ ($_v$=1, 2, 3,...) are the smooth penalized cubic regression spline functions on the climatic indices and the season. MDB~monthlyRain~ is the spatially averaged monthly rainfall across the Murray Darling Basin. Apart from dropping PDO as a predictor, all other climate indices were included, allowing the model to select the appropriate predictors.

@@

One of the difficulties is to point an exact time to the changes in the vegetation cover in the two regions. In the QLD region, no exact time can be assigned to the land clearing. According to the SLATS reports, the most substantial clearing occurred between 2003 - 2004 . However, the information on the change in type of land cover during the time period is missing. Therefore, four scenarios were initially tested in the analysis, however there was no real difference between these scenarios. In the NSW/VIC region, severe bush fires were reported in early January 2003. Hence the "tree" cover state was up to December 2002 then it was changed to "removed" state from January 2003. As a starting date, the regression model was run from 1979 for both regions.

@@

To support the regression analysis, a Mann-Whitney Rank-Sum step trend test was used to detect changes in rainfall as a result of vegetation cover change. This specific nonparametric statistical test was modified from the Mann-Whitney U test by @Hirsch1985 and can identify a step change in data which is cross-correlated. The gridded rainfall dataset used in this study has a high spatial correlation between neighbouring pixels due to the underlying interpolation method. The advantages of using the Rank-Sum test are: (1) it does not depend on assumptions of the data distribution; (2) it is not restricted to datasets with no missing data; (3) it is robust and not as easily influenced by outliers and negative numbers [@Hirsch1985]. However, the test has less power than parametric tests.

@@

```{r exampleRank, echo=F}
table <- data_frame(Year = c(1998 , 1999 , 2000 , 2001 ,
                             2002 , 2005 , 2006),
                    `Rainfall residuals` =c(-0.3, -60.9, -16.1,
                                                      -71.7, 111.1,
                                             -7.2, -60.5),
                    Rank= c(6, 2, 4,
                                                               1, 7, 5, 3))
colnames(table)[3] <- paste("Rank","$R'_{1k}$")
kable(table, caption="Example of ranking rainfall residuals",
             booktabs=T)

@@

The before and after period in the data formed two groups of samples. The split point of the two periods was based on the timing of the vegetation cover changes. In the QLD region, changes occurred anytime during 2003 and 2004. In contrast to the previous method, the time period covering the land cover change  was excluded, as the nonparametric test allows missing data. @Hirsch1985 also pointed out that the power of the test is higher if the data of the change period is ignored. Hence 2003 and 2004 were excluded from the analysis. As a result, the post change period was 2005 - 2015 for the Queensland location.

In the case of NSW/VIC, the bushfires broke out in early January 2003. The change was within a relatively short period of the year. Therefore the post change period in this region still started in January 2003. Following @Hirsch1985, the pre change period was set to five years (1998 - 2002) in both regions. The length of the post change period is also difficult for the NSW/VIC region, as the regrowth would at some point have impacted the local effects.

@@

Instead of completing the Mann-Whitney U-test, @Hirsch1985 applied a standard normal Z test to the rank-sum statistics. As highlighted, this modified test accounts for serial temporal autocorrelation and spatial cross correlation in the data. In the case here, the deseasonalised and detrended data shows little autocorrelation in the time series but possesses strong cross correlation between neighbouring pixels, i.e. $R>0.99$.

@@

```{r Zscore, echo=F}
table2 <- data_frame(" "= c("$Z'>0$  and rainfall decreases after change",
    "$Z'<0$ and rainfall increases post change",
    "$Z'=0$ and rainfall does not change"))
kable(table2, caption="The interpretation of Z' score in the step trend test",
      booktabs=T)
```

@@

```{r rsqmodel, fig.cap = "The distribution of the variance explained (adjusted $r^2$ ) by the full regression model", echo=F, out.width="90%"}
include_graphics("figures/rsq_fullmodel.jpg")

@@

Generally the regression model only explains a limited amount of the rainfall variability (Figure \@reg(fig:rsqmodel)). The model in Equation \@ref(eq:model2) accounts for an average around 33\% of the variation in the NSW/VIC region and about 50% of the variation in the QLD region as indicated by the adjusted $r^2$. The adjusted $r^2$ is the coefficient of determination, a measurement of the amount of variability predicted by the model adjusting for the number of explanatory terms. The residual analysis shows that the assumptions of the regression model are generally met (Figure \@ref(fig:residuals)). The standardised residual plots, however, show some funnelling for the NSW/VIC regions, suggesting non constant variance. The residual patterns are consistent for all pixels within each region.

@@

```{r rsq, fig.cap = "The performance of the regression model if rainfall is only modelled by the climate drivers. It shows the percentage of rainfall variability that can be explained by the climate drivers for the Qld and NSW/VIC region", echo=F, out.width="70%", out.height="70%"}
include_graphics("figures/soi_explhistogram.jpg")

@@

The spatial step trend test based on the regression residuals without the landcover variable generates Z' scores. The results for the Qld region (left) and NSWVIC region (right)  are shown in Figure \@ref(fig:steptest30). This figure provides two types of information: the sign and the significance level. The sign indicates the direction of the step change in the residuals of the regression, as listed in Table \@ref(tab:Zscore). In both regions there are areas of positive Z' values which imply a decrease in rainfall, but this is stronger in the NSW/VIC area than in the QLD area. For the NSW/VIC area there appears to be a reasonable relationship between the locations where changes in tree cover are observed (Figure \@ref(fig:tctrend)) and the patterns in the Z' score. However, there is not necessarily a direct relationship as movement of air masses could mean that actual changes of rainfall are observed close by, but not necessarily exactly at areas with changes of landcover.
There is once again a difference between the left panel and the right panel, indicating only a few significant of Z' scores in the QLD region wich matches the earlier significance in the landcover variable in the full regresion. In the QLD region, only 0.2\% of the pixels obtained a positive Z' score with p $<$ 0.1. In the NSW/VIC region 3.7\% of pixels have a positive Z' score with p $<$ 0.1. In general it is only a small proportion of both study regions.

```{r meandiff, fig.cap ="Boxplots of annual rainfall residuals (estimated based on Equation 2 before and after the land cover intervention during 1979 - 2015 in the study regions. On average, the after period has a significantly lower annual rainfall residual in NSW/VIC, but a signifcantly higher annual rainfall residual in the Qld study area", echo=F, out.width="70%", out.height="70%"}

@@

The rainfall regression residuals without the landcover variable for the two regions (before-change since 1979 and after-change) were also compared (Figure \@ref(fig:meandiff)) using a simple t-test. From the boxplots it is difficult to see that there were significantly different ($p < 0.05$) mean rainfall residual values between the "before" and "after" periods in both regions.  For the Queensland locations, the mean monthly rainfall residuals were slightly lower  ($p < 0.05$) after the change in landcover ($~ 0.45 mm/month$). For the NSW/VIC locations there was larger decrease in the mean monthly rainfall residual ($p < 0.05$) post change ($~ 1.5 mm/month$).
